---
title: "cBioPortal Meta-Analysis Pipeline Summary"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
   rmd: "summary"
output:
  html_document:
    df_print: kable
    toc: true
    toc_depth: 2
---

# Setup

```{r, message=FALSE}
library(tidyverse)
library(arrow)

snek <- snakemake

# settings
MIN_OBS_RATIO <- snek@config$summary$min_gene_coverage_ratio
MIN_SPECIFIC_OBS_RATIO <- snek@config$summary$min_gene_coverage_ratio_specific
NUM_CANCER_SPECIFIC_RESULTS <- snek@config$summary$num_cancer_specific_results
PREVIEW_NUM_ROWS <- snek@config$summary$table_max_rows
SPECIFIC_PREVIEW_NUM_ROWS <- PREVIEW_NUM_ROWS / 5
```

# Studies

```{r studies}
studies <- read_feather(snek@input[[1]])

studies %>%
  arrange(-num_mutations)
```

```{r}
protein_lengths <- read_feather(snek@input[[2]])
```

# Cancer types

## Most common cancer types across studies

```{r cancer_types}
cancers <- read_feather(snek@input[[3]]) %>%
  column_to_rownames("cancer_type") %>%
  select(-mean, -sum)

study_counts <- apply(cancers, 1, function(x) {
  sum(!is.na(x))
})

study_counts <- study_counts %>%
  enframe() %>%
  rename(cancer=name, num_studies=value) %>%
  arrange(-num_studies)

study_counts %>%
  head(PREVIEW_NUM_ROWS)
```

## Cancers with the largest number of mutations across all studies

```{r}
mutation_counts <- rowSums(cancers, na.rm=TRUE)

mutation_counts <- mutation_counts %>%
  enframe() %>%
  rename(cancer=name, num_mutations=value) %>%
  arrange(-num_mutations) %>%
  inner_join(study_counts)

mutation_counts %>%
  head(PREVIEW_NUM_ROWS)
```

# Genes with the largest total number of mutations (pan-cancer)

Which genes have the largest number of mutations across all datasets and patients?

```{r genes}
gene_counts <- read_feather(snek@input[[4]]) %>%
  select(symbol, sum) %>%
  arrange(-sum)

gene_ratios <- read_feather(snek@input[[5]]) %>%
  column_to_rownames("symbol")

num_studies <- ncol(gene_ratios) - 2

gene_ratios$num_nonna <- apply(gene_ratios[, 1:num_studies], 1, function(x) {
  sum(!is.na(x))
})

gene_ratios$ratio_nonna <- gene_ratios$num_nonna / num_studies

num_total <- nrow(gene_ratios)
num_filtered <- sum(gene_ratios$ratio_nonna < MIN_OBS_RATIO)

gene_ratios <- gene_ratios %>%
  filter(ratio_nonna >= MIN_OBS_RATIO)

gene_counts <- gene_counts %>%
  filter(symbol %in% rownames(gene_ratios))
```

```{r results='asis'}
cat(sprintf("Excluded %d / %d genes with mutations observed in fewer than %d / %d studies.\n",
            num_filtered, num_total, ceiling(MIN_OBS_RATIO * num_studies), num_studies))
```


## Overall

```{r gene_mutation_counts_overall}
# add rankings for all genes passing
gene_counts$rank <- seq_len(nrow(gene_counts))

# add protein lengths
gene_counts$protein_length <- round(protein_lengths$length[match(gene_counts$symbol, protein_lengths$symbol)])

median_length <- median(gene_counts$protein_length, na.rm=TRUE)
gene_counts$protein_length[is.na(gene_counts$protein_length)] <- median_length

gene_counts <- gene_counts %>%
  mutate(num_mutations_adj = sum / protein_length) %>%
  select(rank,  symbol, num_mutations=sum, num_mutations_adj, protein_length)

gene_counts %>%
  head(PREVIEW_NUM_ROWS)
```

## Adjusting for protein length

Which genes have the largest number of mutation per amino acid across all datasets and patients?

```{r gene_mutation_counts_adj}
# add rankings for all genes passing
gene_counts <- gene_counts %>%
  arrange(-num_mutations_adj)

gene_counts$rank <- seq_len(nrow(gene_counts))

gene_counts %>%
  head(PREVIEW_NUM_ROWS)
```

# Genes mutated in the largest number of patients (pan-cancer)

## Overall

Which genes are mutated at least once across the largest number of patients?

```{r gene_mutations}
# add rankings for all genes passing
gene_ratios$rank <- seq_len(nrow(gene_ratios))

# add protein lengths
gene_ratios$protein_length <- round(protein_lengths$length[match(rownames(gene_ratios), protein_lengths$symbol)])

median_length <- median(gene_ratios$protein_length, na.rm=TRUE)
gene_ratios$protein_length[is.na(gene_ratios$protein_length)] <- median_length

gene_ratios %>%
  select(rank,  mean, mean_adj, ratio_nonna, num_nonna, protein_length) %>%
  head(PREVIEW_NUM_ROWS)
```

## Adjusting for protein length

Same as above, but dividing by average protein length

```{r gene_mutations_adj}
gene_ratios <- gene_ratios %>%
  arrange(-mean_adj)

gene_ratios$rank_adj <- seq_len(nrow(gene_ratios))

gene_ratios %>%
  select(rank_adj, rank, mean, mean_adj, ratio_nonna, num_nonna, protein_length) %>%
  head(PREVIEW_NUM_ROWS)
```

# Genes mutated in the largest number of cancers (pan-cancer)

## By average scaled mutation count

```{r pan_cancer_gene_rankings1}
gene_cancer_mat <- read_feather(snek@input[[6]]) %>%
  filter(symbol %in% rownames(gene_ratios)) %>%
  column_to_rownames("symbol")

gene_medians <- apply(scale(gene_cancer_mat), 1, median, na.rm=TRUE)

pan_cancer_genes1 <- gene_medians %>%
  enframe(name="symbol", value="median_scaled_mut_count") %>%
  arrange(-median_scaled_mut_count)

pan_cancer_genes1 %>%
  head(PREVIEW_NUM_ROWS)
```

## By average ranking

```{r pan_cancer_gene_rankings2}
# compute gene mutation rankings for each cancer type
cancer_gene_ranks <- data.frame(do.call(cbind, lapply(-gene_cancer_mat, rank)))
rownames(cancer_gene_ranks) <- rownames(gene_cancer_mat)

pan_cancer_genes2 <- apply(cancer_gene_ranks, 1, median, na.rm=TRUE) %>%
  enframe(name="symbol", value="median_mut_rank") %>%
  arrange(median_mut_rank)

pan_cancer_genes2 %>%
  head(PREVIEW_NUM_ROWS)
```

## By average rankings (adjusted)

```{r pan_cancer_gene_rankings3}
# adjust for length
lengths <-  protein_lengths$length[match(rownames(gene_cancer_mat), protein_lengths$symbol)]
lengths[is.na(lengths)] <- median_length

gene_cancer_mat_adj <- sweep(gene_cancer_mat, 1, lengths, "/") * mean(lengths)

gene_medians_adj <- apply(scale(gene_cancer_mat_adj), 1, median, na.rm=TRUE)

pan_cancer_genes3 <- gene_medians_adj %>%
  enframe(name="symbol", value="median_scaled_mut_count_adj")

pan_cancer_genes3$length <- lengths

pan_cancer_genes3 <- pan_cancer_genes3 %>%
  arrange(-median_scaled_mut_count_adj)

pan_cancer_genes3 %>%
  head(PREVIEW_NUM_ROWS)
```

# Genes with the largest total number of mutations (cancer-specific)

```{r gene_mutations_by_cancer_type, results='asis'}
gene_cancer_counts <- read_feather(snek@input[[7]])

# add rankings for all genes passing
common_cancers <- head(study_counts$cancer, NUM_CANCER_SPECIFIC_RESULTS)

for (cancer_ in common_cancers) {
  mask <- as.vector(!is.na(cancers[cancer_, ]))
  studies_with_cancer <- colnames(cancers)[mask]

  df <- gene_cancer_counts %>%
    filter(cancer_type == cancer_) %>%
    select(-cancer_type, -mean, -mean_adj) %>%
    column_to_rownames("symbol")

  df <- df[, studies_with_cancer]

  df$num_nonna <- apply(df, 1, function(x) {
    sum(!is.na(x))
  })

  df$ratio_nonna <- df$num_nonna / length(studies_with_cancer)

  # drop filtered genes
  df <- df %>%
    filter(ratio_nonna >= MIN_SPECIFIC_OBS_RATIO)

  # add protein lengths
  df$protein_length <- round(protein_lengths$length[match(rownames(df), protein_lengths$symbol)])
  df$protein_length[is.na(df$protein_length)] <- median_length

  # compute stats for cancer subset
  df$mean <- rowMeans(df[, studies_with_cancer], na.rm=TRUE)
  df$mean_adj <- df$mean / df$protein_length

  df <- df %>%
    arrange(-mean)

  df$rank <- seq_len(nrow(df))

  # 1. unadjusted
  cat(sprintf("## %s\n\n", cancer_))

  cat("### Overall\n\n")

  df <- df %>%
    select(rank,  mean, mean_adj, ratio_nonna, num_nonna, protein_length)

  df %>%
    head(SPECIFIC_PREVIEW_NUM_ROWS) %>%
    knitr::kable() %>%
    print()

  cat("\n")

  # 2. length-adjusted
  cat("### Adjusted\n\n")

  df <- df %>%
    arrange(-mean_adj)

  df$rank <- seq_len(nrow(df))

  df %>%
    head(SPECIFIC_PREVIEW_NUM_ROWS) %>%
    knitr::kable() %>%
    print()

  cat("\n")
}
```

# Clustering

## Cancers

```{r cancer_clusters_overview}
cancer_clusters <- read_feather(snek@input[[8]])

cancer_clusters %>%
  count(cluster) %>%
  arrange(-n)
```

```{r cancer_clusters, results='asis'}
for (clust_num in sort(unique(cancer_clusters$cluster))) {
  cat(sprintf("## Cancer Cluster %d\n\n", clust_num))

  cat("### Cancers\n\n")

  # print cluster members
  cancer_clusters %>%
    filter(cluster == clust_num) %>%
    select(-cluster) %>%
    knitr::kable() %>%
    print()

  cat("\n")

  # print genes most commonly mutated in this cancer cluster
  cluster_cancers <- cancer_clusters %>%
    filter(cluster == clust_num) %>%
    pull(cancer_type)

  mat <- gene_cancer_mat[, cluster_cancers]

  # drop any cancers with all 0's / zero row variance
  mask <- apply(mat, 2, var) != 0
  mat <- mat[, mask]

  # normalize contribution from each cancer type
  mat <- sweep(mat, 2, colSums(mat), "/")

  # create a dataframe of mutation scaled totals, limited to cancers in cluster
  df <- data.frame(
    symbol = rownames(mat),
    total = rowSums(mat)
  )

  # add protein lengths
  df$protein_length <- round(protein_lengths$length[match(df$symbol, protein_lengths$symbol)])
  df$protein_length[is.na(df$protein_length)] <- median_length

  # add adjusted version of mutation total
  df$adj_total <- df$total / df$protein_length

  df <- df %>%
    arrange(-total) %>%
    mutate(
      rank=dense_rank(desc(total)),
      rank_adj=dense_rank(desc(adj_total))
    ) %>%
    select(rank, rank_adj, total, adj_total, protein_length)

  # 1. unadjusted
  cat("### Genes (unadjusted)\n\n")

  df %>%
    head(PREVIEW_NUM_ROWS) %>%
    knitr::kable() %>%
    print()

  cat("\n")

  # 2. length-adjusted
  cat("### Genes (adjusted)\n\n")

  df <- df %>%
    arrange(-adj_total)

  df %>%
    head(PREVIEW_NUM_ROWS) %>%
    knitr::kable() %>%
    print()

  cat("\n")
}
```

## Genes

```{r gene_clusters_overview}
gene_clusters <- read_feather(snek@input[[9]])

gene_clusters %>%
  count(cluster) %>%
  arrange(-n)
```

```{r gene_clusters, results='asis'}
for (clust_num in sort(unique(gene_clusters$cluster))) {
  cat(sprintf("## Gene Cluster %d\n\n", clust_num))

  cat("### Genes\n\n")

  gene_clusters %>%
    filter(cluster == clust_num) %>%
    select(-cluster) %>%
    head(PREVIEW_NUM_ROWS) %>%
    knitr::kable() %>%
    print()

  cat("\n")
}
```

# Session Info

```{r}
sessionInfo()
```
