---
title: "cBioPortal Meta-Analysis Pipeline Summary"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
   rmd: "summary"
output:
  html_document:
    df_print: kable
    toc: true
    toc_depth: 2
---

# Setup

```{r, message=FALSE}
library(tidyverse)
library(arrow)

snek <- snakemake

# settings
MIN_OBS_RATIO <- snek@config$summary$min_gene_coverage_ratio
NUM_CANCER_SPECIFIC_RESULTS <- snek@config$summary$num_cancer_specific_results
PREVIEW_NUM_ROWS <- snek@config$summary$table_max_rows
```

# Studies

```{r studies}
studies <- read_feather(snek@input[[1]])

studies %>%
  arrange(-num_mutations)
```

# Cancer types

## Most common cancer types across studies

```{r cancer_types}
cancers <- read_feather(snek@input[[2]]) %>%
  column_to_rownames("cancer_type") %>%
  select(-mean, -sum)

study_counts <- apply(cancers, 1, function(x) {
  sum(!is.na(x))
})

study_counts <- study_counts %>%
  enframe() %>%
  rename(cancer=name, num_studies=value) %>%
  arrange(-num_studies)

study_counts %>%
  head(PREVIEW_NUM_ROWS)
```

## Cancers with the largest number of mutations across all studies

```{r}
mutation_counts <- rowSums(cancers, na.rm=TRUE)

mutation_counts <- mutation_counts %>%
  enframe() %>%
  rename(cancer=name, num_mutations=value) %>%
  arrange(-num_mutations) %>%
  inner_join(study_counts)

mutation_counts %>%
  head(PREVIEW_NUM_ROWS)
```

# Genes

```{r genes}
gene_ratios <- read_feather(snek@input[[3]]) %>%
  column_to_rownames("symbol")

protein_lengths <- read_feather(snek@input[[5]])

num_studies <- ncol(gene_ratios) - 2

gene_ratios$num_nonna <- apply(gene_ratios[, 1:num_studies], 1, function(x) {
  sum(!is.na(x))
})

gene_ratios$ratio_nonna <- gene_ratios$num_nonna / num_studies

num_total <- nrow(gene_ratios)
num_filtered <- sum(gene_ratios$ratio_nonna < MIN_OBS_RATIO)

gene_ratios <- gene_ratios %>%
  filter(ratio_nonna >= MIN_OBS_RATIO)
```

```{r results='asis'}
cat(sprintf("Excluded %d / %d genes with mutations observed in fewer than %d / %d studies.\n",
            num_filtered, num_total, ceiling(MIN_OBS_RATIO * num_studies), num_studies))
```

# Gene mutations

## Genes with the highest mutation rates (overall)

```{r gene_mutations}
# add rankings for all genes passing
gene_ratios$rank <- seq_len(nrow(gene_ratios))

# add protein lengths
gene_ratios$protein_length <- round(protein_lengths$length[match(rownames(gene_ratios), protein_lengths$symbol)])

median_length <- median(gene_ratios$protein_length, na.rm=TRUE)
gene_ratios$protein_length[is.na(gene_ratios$protein_length)] <- median_length

gene_ratios %>%
  select(rank,  mean, mean_adj, ratio_nonna, num_nonna, protein_length) %>%
  head(PREVIEW_NUM_ROWS)
```

## Genes with the highest mutation rates (controlling for protein length)

```{r gene_mutations_adj}
gene_ratios <- gene_ratios %>%
  arrange(-mean_adj)

gene_ratios$rank_adj <- seq_len(nrow(gene_ratios))

gene_ratios %>%
  select(rank_adj, rank, mean, mean_adj, ratio_nonna, num_nonna, protein_length) %>%
  head(PREVIEW_NUM_ROWS)
```

# Gene mutations (by cancer type)

## Genes with the highest mutation rates (cancer-specific)

```{r gene_mutations_by_cancer_type, results='asis'}
gene_cancer_ratios <- read_feather(snek@input[[4]])

gene_cancer_ratios$num_nonna <- apply(gene_cancer_ratios[, 3:ncol(gene_cancer_ratios)], 1, function(x) {
  sum(!is.na(x))
})

gene_cancer_ratios$ratio_nonna <- gene_cancer_ratios$num_nonna / num_studies

gene_cancer_ratios <- gene_cancer_ratios %>%
  filter(ratio_nonna >= MIN_OBS_RATIO)

# add rankings for all genes passing
common_cancers <- head(study_counts$cancer, NUM_CANCER_SPECIFIC_RESULTS)

for (cancer in common_cancers) {
  # 1. unadjusted
  cat(sprintf("### %s (overall)\n\n", cancer))

  df <- gene_cancer_ratios %>%
    filter(cancer_type == cancer) %>%
    column_to_rownames("symbol")

  # exclude studies which don't include cancer
  num_nas <- apply(df, 2, function(x) {
    sum(is.na(x))
  })
  mask <- num_nas < nrow(df)

  df <- df[, mask]

  df$rank <- seq_len(nrow(df))

  # add protein lengths
  df$protein_length <- round(protein_lengths$length[match(rownames(df), protein_lengths$symbol)])
  df$protein_length[is.na(df$protein_length)] <- median_length

  df <- df %>%
    select(rank,  mean, mean_adj, ratio_nonna, num_nonna, protein_length)

  df %>%
    head(PREVIEW_NUM_ROWS) %>%
    knitr::kable() %>%
    print()

  cat("\n")

  # 2. length-adjusted
  cat(sprintf("### %s (length-adjusted)\n\n", cancer))

  df <- df %>%
    arrange(-mean_adj)

  df$rank <- seq_len(nrow(df))

  df %>%
    head(PREVIEW_NUM_ROWS) %>%
    knitr::kable() %>%
    print()

  cat("\n")
}
```

# Session Info

```{r}
sessionInfo()
```
